### 2023-12-23 manuscript.tex
initial commit

%%
%% Copyright 2022 OXFORD UNIVERSITY PRESS
%%
%% This file is part of the 'oup-authoring-template Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'oup-authoring-template Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
%% with bibliographic references
%%

%%%CONTEMPORARY%%%
\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
%\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}

%%%MODERN%%%
%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}

%%%TRADITIONAL%%%
%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}

%\onecolumn % for one column layouts

%\usepackage{showframe}

\graphicspath{{Fig/}}

% line numbers
%\usepackage[mathlines, switch]{lineno}
%\usepackage[right]{lineno}

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

\begin{document}

\journaltitle{The ISME Journal}
\DOI{DOI HERE}
\copyrightyear{2024}
\pubyear{2024}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

%\subtitle{Subject Section}

\title[Crete's soil microbiome]{Defining Crete's Macroecology based on its soil Microbial Interactome}

\author[1,2]{Savvas Paragkamian\ORCID{0000-0002-8508-2521}}
\author[3,4]{Johanna B. Holm\ORCID{0000-0002-7646-4085}}
\author[5]{Christina Pavloudi\ORCID{0000-0001-5106-6067}}
\author[6]{Haris Zafeiropoulos\ORCID{0000-0002-4405-6802}}
\author[1]{Christos Christakis\ORCID{0000-0002-7075-0996}}
\author[1]{Melanthia Stavroulaki\ORCID{0000-0003-4392-7159}}
\author[7]{Stephanie A. Yarwood\ORCID{0000-0003-4593-7908}}
\author[2]{Antonios Magoulas}
\author[1]{Panagiotis F. Sarris\ORCID{0000-0001-7000-8997}}
\author[2]{Giorgos Kotoulas\ORCID{0000-0002-4667-8533}}
\author[3]{Lynn Schriml\ORCID{0000-0001-8910-9851}}
\author[2,$\ast$]{Evangelos Pafilis\ORCID{0000-0001-5079-0125}}

\authormark{Paragkamian et al.}

\address[1]{\orgdiv{Department of Biology}, \orgname{University of Crete}, \orgaddress{\street{Voutes}, \postcode{70013}, \state{Crete}, \country{Greece}}}
\address[2]{\orgdiv{Institute of Marine Biology, Biotechnology and Aquaculture (IMBBC)}, \orgname{Hellenic Centre for Marine Research (HCMR)}, \orgaddress{\street{Gournes}, \postcode{71500}, \state{Crete}, \country{Greece}}}
\address[3]{\orgdiv{Institute for Genome Sciences}, \orgname{University of Maryland School of Medicine}, \orgaddress{\state{Baltimore}, \country{USA}}}
\address[4]{\orgdiv{Department of Microbiology and Immunology}, \orgname{University of Maryland School of Medicine}, \orgaddress{\state{Baltimore}, \country{USA}}}
\address[5]{\orgname{European Marine Biological Resource Centre}, \orgaddress{\state{Paris}, \country{France}}}
\address[6]{\orgdiv{Department of Microbiology, Immunology and Transplantation}, \orgname{Rega Institute for Medical Research, KU Leuven}, \orgaddress{\state{Leuven}, \country{Belgium}}}
\address[7]{\orgdiv{Department of Environmental Science and Technology}, \orgname{University of Maryland}, \orgaddress{\state{Baltimore}, \country{USA}}}

\corresp[$\ast$]{Corresponding author. \href{email:pafilis@hcmr.gr}{pafilis@hcmr.gr}}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

%\editor{Associate Editor: Name}

%\abstract{
%\textbf{Motivation:} .\\
%\textbf{Results:} .\\
%\textbf{Availability:} .\\
%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
%online.}

\abstract{Ecosystem functioning is an integral part of the sciences of climate
    change and conservation. Microbes are known for their versatility, abundance
    and influence on ecosystem functions, but a more thorough analysis of
    ecosystems should include a more complex assembly of organisms.
    For example, plant-arthropod-soil microbiome interactions are important
    associations that have historically been ignored. A synthesized knowledge
    base of biodiversity, in terms of ecological and remote-sensing data remains a
    major challenge. Many worldwide studies have been published regarding soil
    microbiome ecosystems, though there are many gaps to cover the complexity
    of functioning and biodiversity in the local scale. Islands can be
    important case studies for this integration, from micro to macro, with Crete being a great example. 
    Here, we utilize the Island Sampling Day Crete 2016 microbial 16S rRNA gene
    amplicon data and metadata, integrated with soil, faunistic and remote
    sensing data, to decipher the drivers of ecosystem function of the island.
    Cretan macroecology has been studied for centuries for its diverse and
    unique geology, fauna and flora. In addition, Crete as a continental island,
    presents a distinct natural and evolutionary history with high contrasts in
    vegetation cover, climatic conditions and geology. The Island Sampling Day
    Crete 2016 project (co-hosted by the Genomic Standards Consortium (GSC) and
    the Institute of Marine Biology, Biotechnology and Aquaculture (IMBBC-HCMR)), has
    collected microbial 16s amplicon data from 72 distinct, with ecosystem
    diversity topsoil sites from all around Crete, accompanied by FAIR
    (Findable, Accessible, Interoperable and Reproducible) data by design. 
    With this island wide study the GSC put the genomic standards in action.
The preliminary results indicate a notable influence of the pH and the elevation
over the island's microbiota. In particular, sites in higher altitudes found to
be inhabited by a more diverse number of microorganisms, a pattern commonly
seen in several faunistic groups, such as arthropods. These data along with the
open access data regarding arthropod fauna, flora and the vast data of remote
sensing and biodiversity, provide the basis to identify major drivers of
biodiversity, to evaluate hotspots and contribute to foreknowledge of threatened ecosystems.}

\keywords{soil microbiome, island biogeography, Crete, data integration, FAIR data}

% \boxedtext{
% \begin{itemize}
% \item Key boxed text here.
% \item Key boxed text here.
% \item Key boxed text here.
% \end{itemize}}

\maketitle

\section{Introduction}\label{intro}

Crete has been studied since the classical times for its' fauna \citep{Sidiropoulos_Polymeni_Legakis_2017}, flora \citep{Krimbas_2005} and environment in general.

The introduction introduces the context and summarizes the manuscript. It is importantly to clearly state the contributions
of this piece of work. Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

This is an example of a new parapgraph with a numbered footnote\footnote{\url{https://data.gov.uk/}} and a second footnote marker.\footnote{Example of footnote text.}

\section{Materials and Methods}\label{methods}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\subsection{Island Sampling Day: Crete}\label{subsec1}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\subsubsection{This is an example for third level head - subsubsection head}\label{subsubsec1}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. %Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\paragraph{This is an example for fourth level head - paragraph head}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\section{Results}\label{results}

\subsection{This is an example for second level head - subsection head}\label{subsec2}

\subsubsection{This is an example for third level head - subsubsection head}\label{subsubsec2}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\paragraph{This is an example for fourth level head - paragraph head}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\section{Equations}\label{sec4}

Equations in \LaTeX{} can either be inline or set as display equations. For
inline equations use the \verb+$...$+ commands. Eg: the equation
$H\psi = E \psi$ is written via the command \verb+$H \psi = E \psi$+.

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum. Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum. 
Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.

\section{Tables}\label{sec5}

Tables can be inserted via the normal table and tabular environment. To put
footnotes inside tables one has to Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum. use the additional ``tablenotes" environment
enclosing the tabular environment. The footnote appears just below the table
itself (refer Tables~\ref{tab1} and \ref{tab2}).

Lengthy tables which do not fit within textwidth should be set as rotated tables. For this, we need to use \verb+\begin{sidewaystable}...+ \verb+\end{sidewaystable}+ instead of\break \verb+\begin{table}...+ \verb+\end{table}+ environment.

\begin{table}[!t]
\caption{Caption text\label{tab1}}%
\begin{tabular*}{\columnwidth}{@{\extracolsep\fill}llll@{\extracolsep\fill}}
\toprule
column 1 & column 2  & column 3 & column 4\\
\midrule
row 1    & data 1   & data 2  & data 3  \\
row 2    & data 4   & data 5$^{1}$  & data 6  \\
row 3    & data 7   & data 8  & data 9$^{2}$  \\
\botrule
\end{tabular*}
\begin{tablenotes}%
\item Source: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote
\item[$^{1}$] Example for a first table footnote.
\item[$^{2}$] Example for a second table footnote.
\end{tablenotes}
\end{table}

\begin{table*}[t]
\caption{Example of a lengthy table which is set to full textwidth.\label{tab2}}
\tabcolsep=0pt%%
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccccc@{\extracolsep{\fill}}}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1$^{1}$} & \multicolumn{3}{@{}c@{}}{Element 2$^{2}$} \\
\cline{2-4}\cline{5-7}%
Project & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3  & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$\\
Element 4  & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$\\
\botrule
\end{tabular*}
\begin{tablenotes}%
\item Note: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote
\item[$^{1}$] Example for a first table footnote.
\item[$^{2}$] Example for a second table footnote.\vspace*{6pt}
\end{tablenotes}
\end{table*}

\section{Figures}\label{sec6}

As per display \LaTeX\ standards one has to use eps images for \verb+latex+ compilation and \verb+pdf/jpg/png+ images for
\verb+pdflatex+ compilation. This is one of the major differences between \verb+latex+
and \verb+pdflatex+. The images should be single-page documents. The command for inserting images
for \verb+latex+ and \verb+pdflatex+ can be generalized. The package used to insert images in \verb+latex/pdflatex+ is the
graphicx package. Figures can be inserted via the normal figure environment as shown in the below example:

\begin{figure}[!t]%
\centering
{\color{black!20}\rule{213pt}{37pt}}
\caption{This is a widefig. This is an example of a long caption this is an example of a long caption  this is an example of a long caption this is an example of a long caption}\label{fig1}
\end{figure}

\begin{figure*}[!t]%
\centering
{\color{black!20}\rule{438pt}{74pt}}
\caption{This is a widefig. This is an example of a long caption this is an example of a long caption  this is an example of a long caption this is an example of a long caption}\label{fig2}
\end{figure*}

For sample purposes, we have included the width of images in the
optional argument of \verb+\includegraphics+ tag. Please ignore this.
Lengthy figures which do not fit within textwidth should be set in rotated mode. For rotated figures, we need to use \verb+\begin{sidewaysfigure}+ \verb+...+ \verb+\end{sidewaysfigure}+ instead of the \verb+\begin{figure}+ \verb+...+ \verb+\end{figure}+ environment.

With standard numerical .bst files, only numerical citations are possible.
With an author-year .bst file, both numerical and author-year citations are possible.

\noindent
Multiple citations as normal:

\section{Discussion}\label{discussion}

\section{Conclusion}

Some Conclusions here.

%%%%%%%%%%%%%%

%\begin{appendices}
%
%\section{Appendix 1}\label{appendix1}
%
%Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
%pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
%vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
%a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
%urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris.
%
%
%\section{Appendix 2}\label{appendix2}
%\subsubsection{Subsubsection title of first appendix}\label{subsubsec3}
%
%Fusce mauris. Vestibulum luctus nibh at lectus. Sed bibendum, nulla a faucibus semper, leo velit ultricies tellus, ac
%venenatis arcu wisi vel nisl. Vestibulum diam. Aliquam pellentesque, augue quis sagittis posuere, turpis lacus congue
%quam, in hendrerit risus eros eget felis.
%
%\end{appendices}

\section{Competing interests}
No competing interest is declared.

\section{Author contributions statement}

Must include all authors, identified by initials, for example:
S.R. and D.A. conceived the experiment(s),  S.R. conducted the experiment(s), S.R. and D.A. analysed the results.  S.R. and D.A. wrote and reviewed the manuscript.

\section{Acknowledgments}
The authors thank the anonymous reviewers for their valuable suggestions. This work is supported in part by funds from the National Science Foundation (NSF: \# 1636933 and \# 1920920).

\bibliographystyle{plain}
\bibliography{reference}

%USE THE BELOW OPTIONS IN CASE YOU NEED AUTHOR YEAR FORMAT.
%\bibliographystyle{abbrvnat}
%\bibliography{reference}

\end{document}

### 2023-12-23 manuscript.tex
add ref

@@
 
-Crete has been studied since the classical times for its' fauna \citep{Sidiropoulos_Polymeni_Legakis_2017}, flora \citep{Krimbas_2005} and environment in general.
+The island of Crete, Greece has been studied since the classical times for its' fauna \citep{Sidiropoulos_Polymeni_Legakis_2017}, flora \citep{Krimbas_2005} and ecosystems \citep{Grove1993}.
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{unsrt}
 \bibliography{reference}

### 2023-12-23 manuscript.tex
add more refs and draft text in intro

@@
 The island of Crete, Greece has been studied since the classical times for its' fauna \citep{Sidiropoulos_Polymeni_Legakis_2017}, flora \citep{Krimbas_2005} and ecosystems \citep{Grove1993}.
+Crete is a continental island, fifth largest island of the mediterranean, a mediterranean biodiversity hotspot, model for desertification and climate change, elevation gradient, grasing impacts on ecosystems and evolutionary model.
+Yet the terrestrial microbial diversity has been unexplored.
 
-The introduction introduces the context and summarizes the manuscript. It is importantly to clearly state the contributions
-of this piece of work.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Has been sampled from major worldwide projects of microbiome studies.
 
-This is an example of a new parapgraph with a numbered footnote\footnote{\url{https://data.gov.uk/}} and a second footnote marker.\footnote{Example of footnote text.}
+Global soil microbiome studies \citep{Delgado-Baquerizo2018} and \citep{Bahram2018} and the biogeography \citep{Martiny2006} \citep{Guerra2020}, \citep{Labouyrie2023} yet there are significant blind spots \citep{Guerra2020}.
+The most resolute study is \citep{Karimi2020}
+
+Islands are nature's labs \citep{Whittaker2017} and in soil microbes \citep{Li2020} and fungi \citep{Delavaux2021}
+Ultimately islands can set the ground to represent ecosystems with data and complex interactions \citep{Davies2016}.
+
+Genome Standards Consortium, FAIR data, unrestricted, rich metadata
+
+Here we present
+
+%This is an example of a new parapgraph with a numbered footnote\footnote{\url{https://data.gov.uk/}} and a second footnote marker.\footnote{Example of footnote text.}
 

### 2023-12-24 manuscript.tex
add points with cited references across the document

@@
 \author[6]{Haris Zafeiropoulos\ORCID{0000-0002-4405-6802}}
-\author[1]{Christos Christakis\ORCID{0000-0002-7075-0996}}
+\author[1]{Christos A. Christakis\ORCID{0000-0002-7075-0996}}
 \author[1]{Melanthia Stavroulaki\ORCID{0000-0003-4392-7159}}
@@
 
-The island of Crete, Greece has been studied since the classical times for its' fauna \citep{Sidiropoulos_Polymeni_Legakis_2017}, flora \citep{Krimbas_2005} and ecosystems \citep{Grove1993}.
-Crete is a continental island, fifth largest island of the mediterranean, a mediterranean biodiversity hotspot, model for desertification and climate change, elevation gradient, grasing impacts on ecosystems and evolutionary model.
+The island of Crete, Greece has been studied since the classical times for its' fauna \citep{Sidiropoulos_Polymeni_Legakis_2017,Anastasiou2018Tenebrionid}, flora \citep{Krimbas_2005} and ecosystems \citep{Grove1993}.
+Crete is a continental island, fifth largest island of the mediterranean, a mediterranean biodiversity hotspot, model for desertification, climate change \citep{Kougioumoutzis2020} and habitat loss \citep{ISPIKOUDIS1993259},
+elevation gradient \citep{trigas2013, FAZAN2017}, grasing impacts on ecosystems \citep{JouffroyBapicot2016} and evolutionary model \citep{POULAKAKIS2002}.
 Yet the terrestrial microbial diversity has been unexplored.
 
-Has been sampled from major worldwide projects of microbiome studies.
+Has been sampled from major worldwide projects of microbiome studies \citep{Vasar2022, Labouyrie2023, Bahram2018, Orgiazzi2018}.
+Soil fungi \citep{Mikryukov2023, Davison2021, Tedersoo2021} and other soil eukaryotes \citep{Aslani2022}
 
@@
 
-Genome Standards Consortium, FAIR data, unrestricted, rich metadata
+Genome Standards Consortium \citep{Field2011}, FAIR data \citep{Wilkinson2016}, unrestricted \citep{Amann2019},
+rich metadata and platforms of hosting open data like European Soil Data Centre \citep{Panagos2022}
 
@@
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+\subsection{Island Sampling Day: Crete}\label{isd_data}
 
-\subsection{Island Sampling Day: Crete}\label{subsec1}
+the data for this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB21776
+ENA API \citep{Yuan2023}
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+\subsection{Data integration}\label{data}
+
+Copernicus CORINE Land Cover \citep{CLC2023}
+WorldClim 2.0 \citep{Fick2017}
+World Database on Protected Areas (WDPA) \citep{Hanson2022}
+Function annotation - FAPROTAX \citep{Louca2016}
+Crete geology from IGME
+Aridity
+%NDVI
+Taxonomy Silva 138 \citep{Quast2012}
+
+\subsection{Tools}\label{tools}
+dada2 for ASV inferrence \citep{Callahan2016}
+PEMA for OTU inferrence \citep{Zafeiropoulos2020}
+ape R packege \citep{Paradis2004}
+SRS R package \citep{Beule2020}
+UMAP ordination \citep{mcinnes2018umap-software}
+Numerical ecology analyses with vegan R package \citep{vegan}
+
+Python 3.11.4
+Julia Version 1.9.3 \citep{Julia-2017}
+R version 4.3.2 \citep{rcoreteam}
+Network inferrence using FlashWeave 0.19.2 \citep{Tackmann2019}
+Network analysis in igraph R package \citep{Csardi2006}
+Differential abundance analysis with ANCOM-BC R package \citep{Lin2020}
+Spatial analysis with the sf and stars R packages \citep{Pebesma2023}
+U-CIE R package for coloring 3 dimentional data \citep{Koutrouli2022}
+ggplot2 \citep{ggplot22016}
+pheatmap \citep{Kolde2019}
+
+Computations were performed on HPC infrastrure of HCMR \citep{Zafeiropoulos2021}.
 
@@
 Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-
-\section{Equations}\label{sec4}
-
-Equations in \LaTeX{} can either be inline or set as display equations. For
-inline equations use the \verb+$...$+ commands. Eg: the equation
-$H\psi = E \psi$ is written via the command \verb+$H \psi = E \psi$+.
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
 
@@
 
+Networks specialists generalists \citep{Barberan2012}
+
+Elevation gradients of biodiversity are known since Humboldt's work \citep{Rahbek2019} 
+yet these patterns remain elusive regarding the soil microbiome \citep{Looby2020, Siles2023}.
+Mostly because it's difficult to isolate other cofounding effects \citep{Nottingham2018}.
+
+Functions of Richtis gorge are a signal in need to be further investigated.
+
+Enumeratic all biodivarsity \citep{Anthony2023} and decipher the biogeochemical 
+processes and interactions of fauna, flora and microbes \citep{Fry2019, Crowther2019,GRANDY201640,Delgado-Baquerizo2020}
+
+Global hotspots \citep{Guerra2022}
+and soil conservation instead of specific species \citep{Guerra2021}
+and policy \citep{KONINGER2022} across countries \citep{Putten2023}
+
 \section{Conclusion}
 
+Mountain peaks soil microbes \citep{Adamczyk2019}
+These studies can act as early warning signals towards public health conserns \citep{Banerjee2023}.
 Some Conclusions here.

### 2023-12-25 manuscript.tex
add refs of Vogiatzakis

@@
 
-\title[Crete's soil microbiome]{Defining Crete's Macroecology based on its soil Microbial Interactome}
+\title[Crete's soil microbiome]{Exploring Crete's Macroecology based on its soil Microbial Interactome}
 
@@
 Ultimately islands can set the ground to represent ecosystems with data and complex interactions \citep{Davies2016}.
+\citep{Vogiatzakis2008_crete}
+\citep{Vogiatzakis2016}
 
+\citep{Sfenthourakis2017}
+\citep{Vogiatzakis2008_med}
 Genome Standards Consortium \citep{Field2011}, FAIR data \citep{Wilkinson2016}, unrestricted \citep{Amann2019},

### 2023-12-30 manuscript.tex
restrure the flow of intro

@@
 
+Soil ecosystems are the cornstone of terrestrial habitats, biodiversity and henceforth human activities.
+Complex interactions of multiple systems, nutrient cycling, atmosphere, climate, biogeochemical cycles and biodiversity 
+across all forms of life, fauna, flora, bacteria, archaea, fungi, viruses. 
+Bacteria and archaea are considered major drivers for the functionality of soil.
+They infuence and are infuenced by their environments and their community structure 
+is defies their macroscopic functionality.
+Global soil microbiome studies \citep{Delgado-Baquerizo2018} and \citep{Bahram2018} and the biogeography \citep{Martiny2006} \citep{Guerra2020}, \citep{Labouyrie2023} yet there are significant blind spots \citep{Guerra2020}.
+The most resolute study is \citep{Karimi2020} which exemplified the
+remarkable complexity of soil bacterial communuties and the requirement of  dense samplings and isolated systems.
+
+Island Sampling Day project and the 
+Genome Standards Consortium \citep{Field2011} "put standards into action" and the 
+18th workshop. 
+The choice of an island
+Islands are nature's labs \citep{Whittaker2017} and in soil microbes \citep{Li2020} and fungi \citep{Delavaux2021}
+Ultimately islands can set the ground to represent ecosystems with data and complex interactions \citep{Davies2016}.
+A large scale study in the confined space of the island of Crete, a miniature continent.
+
 The island of Crete, Greece has been studied since the classical times for its' fauna \citep{Sidiropoulos_Polymeni_Legakis_2017,Anastasiou2018Tenebrionid}, flora \citep{Krimbas_2005} and ecosystems \citep{Grove1993}.
@@
 
-Global soil microbiome studies \citep{Delgado-Baquerizo2018} and \citep{Bahram2018} and the biogeography \citep{Martiny2006} \citep{Guerra2020}, \citep{Labouyrie2023} yet there are significant blind spots \citep{Guerra2020}.
-The most resolute study is \citep{Karimi2020}
-
-Islands are nature's labs \citep{Whittaker2017} and in soil microbes \citep{Li2020} and fungi \citep{Delavaux2021}
-Ultimately islands can set the ground to represent ecosystems with data and complex interactions \citep{Davies2016}.
 \citep{Vogiatzakis2008_crete}
@@
 \citep{Vogiatzakis2008_med}
+
+Integration of prior knowledge and 
 Genome Standards Consortium \citep{Field2011}, FAIR data \citep{Wilkinson2016}, unrestricted \citep{Amann2019},
@@
 
-Here we present
+Interactomes as a driver of community composition. Microbial interactions challenges \citep{Faust2021} cohesion?
+
+In this study, we investigate
 

### 2024-01-04 manuscript.tex
add ancombc ref and fix italics

@@
 %\usepackage{showframe}
-
+\usepackage[T1]{fontenc}
 \graphicspath{{Fig/}}
@@
 The island of Crete, Greece has been studied since the classical times for its' fauna \citep{Sidiropoulos_Polymeni_Legakis_2017,Anastasiou2018Tenebrionid}, flora \citep{Krimbas_2005} and ecosystems \citep{Grove1993}.
+Crete is home to the only endemic mammal of Greece, the Cretan shrew (\textit{Crocidura zimmermanni}), more than 350 endemic arhropods, \# endemic plants among them a tree \textit{Zelkova abelicea}.
 Crete is a continental island, fifth largest island of the mediterranean, a mediterranean biodiversity hotspot, model for desertification, climate change \citep{Kougioumoutzis2020} and habitat loss \citep{ISPIKOUDIS1993259},
@@
 Network analysis in igraph R package \citep{Csardi2006}
-Differential abundance analysis with ANCOM-BC R package \citep{Lin2020}
+Differential abundance analysis with ANCOM-BC R package \citep{Lin2020,Lin2023}
 Spatial analysis with the sf and stars R packages \citep{Pebesma2023}

### 2024-02-26 manuscript.tex
add tsaparis as author
add ref for soil A critical perspective on interpreting amplicon sequencing

@@
 \author[1]{Melanthia Stavroulaki\ORCID{0000-0003-4392-7159}}
+\author[1]{Dimitris Tsaparis\ORCID{0000-0001-9626-5553}}
+%\author[1]{Iraklis Vretzakis\ORCID{0000-0001-9626-5553}}
 \author[7]{Stephanie A. Yarwood\ORCID{0000-0003-4593-7908}}
@@
 
-\keywords{soil microbiome, island biogeography, Crete, data integration, FAIR data}
+\keywords{soil microbiome, networks, island biogeography, Crete, data integration, FAIR data}
 
@@
 Mountain peaks soil microbes \citep{Adamczyk2019}
-These studies can act as early warning signals towards public health conserns \citep{Banerjee2023}.
+Even though amplicon studies in soil should be interpeted with caution \citep{alteio2021} they 
+can act as early warning signals towards public health conserns \citep{Banerjee2023}.
 Some Conclusions here.

### 2024-02-26 manuscript.tex
move to cite from citep

@@
 is defies their macroscopic functionality.
-Global soil microbiome studies \citep{Delgado-Baquerizo2018} and \citep{Bahram2018} and the biogeography \citep{Martiny2006} \citep{Guerra2020}, \citep{Labouyrie2023} yet there are significant blind spots \citep{Guerra2020}.
-The most resolute study is \citep{Karimi2020} which exemplified the
+Global soil microbiome studies \cite{Delgado-Baquerizo2018} and \cite{Bahram2018} and the biogeography \cite{Martiny2006} \cite{Guerra2020}, \cite{Labouyrie2023} yet there are significant blind spots \cite{Guerra2020}.
+The most resolute study is \cite{Karimi2020} which exemplified the
 remarkable complexity of soil bacterial communuties and the requirement of dense samplings and isolated systems.
@@
 Island Sampling Day project and the 
-Genome Standards Consortium \citep{Field2011} "put standards into action" and the 
+Genome Standards Consortium \cite{Field2011} "put standards into action" and the 
 18th workshop. 
 The choice of an island
-Islands are nature's labs \citep{Whittaker2017} and in soil microbes \citep{Li2020} and fungi \citep{Delavaux2021}
-Ultimately islands can set the ground to represent ecosystems with data and complex interactions \citep{Davies2016}.
+Islands are nature's labs \cite{Whittaker2017} and in soil microbes \cite{Li2020} and fungi \cite{Delavaux2021}
+Ultimately islands can set the ground to represent ecosystems with data and complex interactions \cite{Davies2016}.
 A large scale study in the confined space of the island of Crete, a miniature continent.
 
-The island of Crete, Greece has been studied since the classical times for its' fauna \citep{Sidiropoulos_Polymeni_Legakis_2017,Anastasiou2018Tenebrionid}, flora \citep{Krimbas_2005} and ecosystems \citep{Grove1993}.
+The island of Crete, Greece has been studied since the classical times for its' fauna \cite{Sidiropoulos_Polymeni_Legakis_2017,Anastasiou2018Tenebrionid}, flora \cite{Krimbas_2005} and ecosystems \cite{Grove1993}.
 Crete is home to the only endemic mammal of Greece, the Cretan shrew (\textit{Crocidura zimmermanni}), more than 350 endemic arhropods, \# endemic plants among them a tree \textit{Zelkova abelicea}.
-Crete is a continental island, fifth largest island of the mediterranean, a mediterranean biodiversity hotspot, model for desertification, climate change \citep{Kougioumoutzis2020} and habitat loss \citep{ISPIKOUDIS1993259},
-elevation gradient \citep{trigas2013, FAZAN2017}, grasing impacts on ecosystems \citep{JouffroyBapicot2016} and evolutionary model \citep{POULAKAKIS2002}.
+Crete is a continental island, fifth largest island of the mediterranean, a mediterranean biodiversity hotspot, model for desertification, climate change \cite{Kougioumoutzis2020} and habitat loss \cite{ISPIKOUDIS1993259},
+elevation gradient \cite{trigas2013, FAZAN2017}, grasing impacts on ecosystems \cite{JouffroyBapicot2016} and evolutionary model \cite{POULAKAKIS2002}.
 Yet the terrestrial microbial diversity has been unexplored.
 
-Has been sampled from major worldwide projects of microbiome studies \citep{Vasar2022, Labouyrie2023, Bahram2018, Orgiazzi2018}.
-Soil fungi \citep{Mikryukov2023, Davison2021, Tedersoo2021} and other soil eukaryotes \citep{Aslani2022}
+Has been sampled from major worldwide projects of microbiome studies \cite{Vasar2022, Labouyrie2023, Bahram2018, Orgiazzi2018}.
+Soil fungi \cite{Mikryukov2023, Davison2021, Tedersoo2021} and other soil eukaryotes \cite{Aslani2022}
 
-\citep{Vogiatzakis2008_crete}
-\citep{Vogiatzakis2016}
+\cite{Vogiatzakis2008_crete}
+\cite{Vogiatzakis2016}
 
-\citep{Sfenthourakis2017}
-\citep{Vogiatzakis2008_med}
+\cite{Sfenthourakis2017}
+\cite{Vogiatzakis2008_med}
 
 Integration of prior knowledge and 
-Genome Standards Consortium \citep{Field2011}, FAIR data \citep{Wilkinson2016}, unrestricted \citep{Amann2019},
-rich metadata and platforms of hosting open data like European Soil Data Centre \citep{Panagos2022}
+Genome Standards Consortium \cite{Field2011}, FAIR data \cite{Wilkinson2016}, unrestricted \cite{Amann2019},
+rich metadata and platforms of hosting open data like European Soil Data Centre \cite{Panagos2022}
 
-Interactomes as a driver of community composition. Microbial interactions challenges \citep{Faust2021} cohesion?
+Interactomes as a driver of community composition. Microbial interactions challenges \cite{Faust2021} cohesion?
 
@@
 the data for this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB21776
-ENA API \citep{Yuan2023}
+ENA API \cite{Yuan2023}
 
@@
 
-Copernicus CORINE Land Cover \citep{CLC2023}
-WorldClim 2.0 \citep{Fick2017}
-World Database on Protected Areas (WDPA) \citep{Hanson2022}
-Function annotation - FAPROTAX \citep{Louca2016}
+Copernicus CORINE Land Cover \cite{CLC2023}
+WorldClim 2.0 \cite{Fick2017}
+World Database on Protected Areas (WDPA) \cite{Hanson2022}
+Function annotation - FAPROTAX \cite{Louca2016}
 Crete geology from IGME
@@
 %NDVI
-Taxonomy Silva 138 \citep{Quast2012}
+Taxonomy Silva 138 \cite{Quast2012}
 
 \subsection{Tools}\label{tools}
-dada2 for ASV inferrence \citep{Callahan2016}
-PEMA for OTU inferrence \citep{Zafeiropoulos2020}
-ape R packege \citep{Paradis2004}
-SRS R package \citep{Beule2020}
-UMAP ordination \citep{mcinnes2018umap-software}
-Numerical ecology analyses with vegan R package \citep{vegan}
+dada2 for ASV inferrence \cite{Callahan2016}
+PEMA for OTU inferrence \cite{Zafeiropoulos2020}
+ape R packege \cite{Paradis2004}
+SRS R package \cite{Beule2020}
+UMAP ordination \cite{mcinnes2018umap-software}
+Numerical ecology analyses with vegan R package \cite{vegan}
 
 Python 3.11.4
-Julia Version 1.9.3 \citep{Julia-2017}
-R version 4.3.2 \citep{rcoreteam}
-Network inferrence using FlashWeave 0.19.2 \citep{Tackmann2019}
-Network analysis in igraph R package \citep{Csardi2006}
-Differential abundance analysis with ANCOM-BC R package \citep{Lin2020,Lin2023}
-Spatial analysis with the sf and stars R packages \citep{Pebesma2023}
-U-CIE R package for coloring 3 dimentional data \citep{Koutrouli2022}
-ggplot2 \citep{ggplot22016}
-pheatmap \citep{Kolde2019}
+Julia Version 1.9.3 \cite{Julia-2017}
+R version 4.3.2 \cite{rcoreteam}
+Network inferrence using FlashWeave 0.19.2 \cite{Tackmann2019}
+Network analysis in igraph R package \cite{Csardi2006}
+Differential abundance analysis with ANCOM-BC R package \cite{Lin2020,Lin2023}
+Spatial analysis with the sf and stars R packages \cite{Pebesma2023}
+U-CIE R package for coloring 3 dimentional data \cite{Koutrouli2022}
+ggplot2 \cite{ggplot22016}
+pheatmap \cite{Kolde2019}
 
-Computations were performed on HPC infrastrure of HCMR \citep{Zafeiropoulos2021}.
+Computations were performed on HPC infrastrure of HCMR \cite{Zafeiropoulos2021}.
 
@@
 
-Networks specialists generalists \citep{Barberan2012}
+Networks specialists generalists \cite{Barberan2012}
 
-Elevation gradients of biodiversity are known since Humboldt's work \citep{Rahbek2019} 
-yet these patterns remain elusive regarding the soil microbiome \citep{Looby2020, Siles2023}.
-Mostly because it's difficult to isolate other cofounding effects \citep{Nottingham2018}.
+Elevation gradients of biodiversity are known since Humboldt's work \cite{Rahbek2019} 
+yet these patterns remain elusive regarding the soil microbiome \cite{Looby2020, Siles2023}.
+Mostly because it's difficult to isolate other cofounding effects \cite{Nottingham2018}.
 
@@
 
-Enumeratic all biodivarsity \citep{Anthony2023} and decipher the biogeochemical 
-processes and interactions of fauna, flora and microbes \citep{Fry2019, Crowther2019,GRANDY201640,Delgado-Baquerizo2020}
+Enumeratic all biodivarsity \cite{Anthony2023} and decipher the biogeochemical 
+processes and interactions of fauna, flora and microbes \cite{Fry2019, Crowther2019,GRANDY201640,Delgado-Baquerizo2020}
 
-Global hotspots \citep{Guerra2022}
-and soil conservation instead of specific species \citep{Guerra2021}
-and policy \citep{KONINGER2022} across countries \citep{Putten2023}
+Global hotspots \cite{Guerra2022}
+and soil conservation instead of specific species \cite{Guerra2021}
+and policy \cite{KONINGER2022} across countries \cite{Putten2023}
 
@@
 
-Mountain peaks soil microbes \citep{Adamczyk2019}
-Even though amplicon studies in soil should be interpeted with caution \citep{alteio2021} they 
-can act as early warning signals towards public health conserns \citep{Banerjee2023}.
+Mountain peaks soil microbes \cite{Adamczyk2019}
+Even though amplicon studies in soil should be interpeted with caution \cite{alteio2021} they 
+can act as early warning signals towards public health conserns \cite{Banerjee2023}.
 Some Conclusions here.

### 2024-04-02 manuscript.tex
structure update

@@
 
-\title[Crete's soil microbiome]{Exploring Crete's Macroecology based on its soil Microbial Interactome}
+\title[Crete's soil microbiome]{Exploring the Soil Microbiome coupling with the Crete island Data Cube}
 
@@
 
-Soil ecosystems are the cornstone of terrestrial habitats, biodiversity and henceforth human activities.
-Complex interactions of multiple systems, nutrient cycling, atmosphere, climate, biogeochemical cycles and biodiversity 
-across all forms of life, fauna, flora, bacteria, archaea, fungi, viruses. 
+Soil ecosystems are the cornerstone of terrestrial habitats, biodiversity and henceforth human activities.
+Soils are characterised by multiple properties; chemical, physical and biological that 
+form complex interdependant interactions. Biodiversity of soils covers
+all forms of life, fauna, flora, bacteria, archaea, fungi, viruses. 
 Bacteria and archaea are considered major drivers for the functionality of soil.
-They infuence and are infuenced by their environments and their community structure 
-is defies their macroscopic functionality.
-Global soil microbiome studies \cite{Delgado-Baquerizo2018} and \cite{Bahram2018} and the biogeography \cite{Martiny2006} \cite{Guerra2020}, \cite{Labouyrie2023} yet there are significant blind spots \cite{Guerra2020}.
-The most resolute study is \cite{Karimi2020} which exemplified the
+They infuence and are infuenced by their environment and their community structure 
+defies their macroscopic functionality.
+Global soil microbiome studies have been employed to decipher soil microbiome compositions \cite{Delgado-Baquerizo2018, Labouyrie2023},
+functions \cite{Bahram2018} and biogeography \cite{Martiny2006, Guerra2020}.
+These results showed the remarkable diversity in soils yet there are blind spots \cite{Guerra2020}
+and these sampling are sparce when considering samples per area density. One of most resolute
+study is by \cite{Karimi2020} which exemplified the 
 remarkable complexity of soil bacterial communuties and the requirement of dense samplings and isolated systems.
 
+Islands are nature's labs \cite{Whittaker2017} because of their isolation and smaller scale.
+Borrowing this paradigm, soil microbes \cite{Li2020} and mycorrhizal fungi \cite{Delavaux2021} studies
+in islands show the benefits of using islands as models.
 Island Sampling Day project and the 
@@
 The choice of an island
-Islands are nature's labs \cite{Whittaker2017} and in soil microbes \cite{Li2020} and fungi \cite{Delavaux2021}
 Ultimately islands can set the ground to represent ecosystems with data and complex interactions \cite{Davies2016}.
@@
 
-%This is an example of a new parapgraph with a numbered footnote\footnote{\url{https://data.gov.uk/}} and a second footnote marker.\footnote{Example of footnote text.}
-
 \section{Materials and Methods}\label{methods}
@@
 
-the data for this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB21776
+The data for this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB21776
 ENA API \cite{Yuan2023}
@@
 
-\subsubsection{This is an example for third level head - subsubsection head}\label{subsubsec1}
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-%Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-
-\paragraph{This is an example for fourth level head - paragraph head}
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-
 \section{Results}\label{results}
 
-\subsection{This is an example for second level head - subsection head}\label{subsec2}
+\subsection{Soil microbiome}\label{soil_microbiome}
 
-\subsubsection{This is an example for third level head - subsubsection head}\label{subsubsec2}
+Summary of the pipeline results
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Generalists and specialists
 
-\paragraph{This is an example for fourth level head - paragraph head}
+\subsection{Crete Data Cube}\label{data_cube}
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+\subsection{Community drivers}\label{Drivers}
 
-\section{Tables}\label{sec5}
+\subsection{Interactome}\label{interactome}
 
-Tables can be inserted via the normal table and tabular environment. To put
-footnotes inside tables one has to Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-use the additional ``tablenotes" environment
-enclosing the tabular environment. The footnote appears just below the table
-itself (refer Tables~\ref{tab1} and \ref{tab2}).
+\subsection{Human interventions}\label{Human interventions}
 
-Lengthy tables which do not fit within textwidth should be set as rotated tables.
-For this, we need to use \verb+\begin{sidewaystable}...+ \verb+\end{sidewaystable}+ instead of\break \verb+\begin{table}...+ \verb+\end{table}+ environment.
+\subsubsection{Functions}\label{functions}
 
-\begin{table}[!t]
-\caption{Caption text\label{tab1}}%
-\begin{tabular*}{\columnwidth}{@{\extracolsep\fill}llll@{\extracolsep\fill}}
-\toprule
-column 1 & column 2  & column 3 & column 4\\
-\midrule
-row 1    & data 1   & data 2  & data 3  \\
-row 2    & data 4   & data 5$^{1}$  & data 6  \\
-row 3    & data 7   & data 8  & data 9$^{2}$  \\
-\botrule
-\end{tabular*}
-\begin{tablenotes}%
-\item Source: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote
-\item[$^{1}$] Example for a first table footnote.
-\item[$^{2}$] Example for a second table footnote.
-\end{tablenotes}
-\end{table}
+\subsubsection{Significant taxa}\label{sig_taxa}
 
-\begin{table*}[t]
-\caption{Example of a lengthy table which is set to full textwidth.\label{tab2}}
-\tabcolsep=0pt%%
-\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccccc@{\extracolsep{\fill}}}
-\toprule%
-& \multicolumn{3}{@{}c@{}}{Element 1$^{1}$} & \multicolumn{3}{@{}c@{}}{Element 2$^{2}$} \\
-\cline{2-4}\cline{5-7}%
-Project & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
-\midrule
-Element 3  & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$\\
-Element 4  & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$\\
-\botrule
-\end{tabular*}
-\begin{tablenotes}%
-\item Note: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote
-\item[$^{1}$] Example for a first table footnote.
-\item[$^{2}$] Example for a second table footnote.\vspace*{6pt}
-\end{tablenotes}
-\end{table*}
-
-\section{Figures}\label{sec6}
-
-As per display \LaTeX\ standards one has to use eps images for \verb+latex+ compilation and \verb+pdf/jpg/png+ images for
-\verb+pdflatex+ compilation. This is one of the major differences between \verb+latex+
-and \verb+pdflatex+. The images should be single-page documents. The command for inserting images
-for \verb+latex+ and \verb+pdflatex+ can be generalized. The package used to insert images in \verb+latex/pdflatex+ is the
-graphicx package. Figures can be inserted via the normal figure environment as shown in the below example:
-
-\begin{figure}[!t]%
-\centering
-{\color{black!20}\rule{213pt}{37pt}}
-\caption{This is a widefig. This is an example of a long caption this is an example of a long caption  this is an example of a long caption this is an example of a long caption}\label{fig1}
-\end{figure}
-
-\begin{figure*}[!t]%
-\centering
-{\color{black!20}\rule{438pt}{74pt}}
-\caption{This is a widefig. This is an example of a long caption this is an example of a long caption  this is an example of a long caption this is an example of a long caption}\label{fig2}
-\end{figure*}
-
-For sample purposes, we have included the width of images in the
-optional argument of \verb+\includegraphics+ tag. Please ignore this.
-Lengthy figures which do not fit within textwidth should be set in rotated mode.
-For rotated figures, we need to use \verb+\begin{sidewaysfigure}+ \verb+...+ \verb+\end{sidewaysfigure}+ instead of the \verb+\begin{figure}+ \verb+...+ \verb+\end{figure}+ environment.
-
-With standard numerical .bst files, only numerical citations are possible.
-With an author-year .bst file, both numerical and author-year citations are possible.
-
-\noindent
-Multiple citations as normal:
+Differential abundance
 
@@
 
+"A holistic perspective on soil architecture is needed as a key to soil functions"
 \section{Conclusion}
@@
 can act as early warning signals towards public health conserns \cite{Banerjee2023}.
-Some Conclusions here.
+
+Future works, shotgun metagenomics, long reads, need for even higher resolution samplings.
 

### 2024-04-03 manuscript.tex
intro half

@@
 study is by \cite{Karimi2020} which exemplified the 
-remarkable complexity of soil bacterial communuties and the requirement of dense samplings and isolated systems.
+vast complexity of soil bacterial communuties and the requirement of dense samplings and isolated systems.
 
@@
 Borrowing this paradigm, soil microbes \cite{Li2020} and mycorrhizal fungi \cite{Delavaux2021} studies
-in islands show the benefits of using islands as models.
-Island Sampling Day project and the 
-Genome Standards Consortium \cite{Field2011} "put standards into action" and the 
-18th workshop. 
-The choice of an island
-Ultimately islands can set the ground to represent ecosystems with data and complex interactions \cite{Davies2016}.
-A large scale study in the confined space of the island of Crete, a miniature continent.
+in islands show the benefits of using islands as models. Ultimately islands canr
+set the ground to represent ecosystems with data and complex interactions \cite{Davies2016}.
+This was also the case for the Island Sampling Day (ISD) project \cite{holm2024}
+of the Genome Standards Consortium \cite{Field2011}
+during the 18th workshop in June 2026 in Crete island, Greece. The goal was to "put standards into action"
+in a soil microbiome survey with a dense sampling, 0.017 samples per km\textsuperscript{2}.
+Hence, ISD is large scale study in the confined space of the island of Crete which 
+is considered a miniature continent \cite{Vogiatzakis2008_crete}.
 
-The island of Crete, Greece has been studied since the classical times for its' fauna \cite{Sidiropoulos_Polymeni_Legakis_2017,Anastasiou2018Tenebrionid}, flora \cite{Krimbas_2005} and ecosystems \cite{Grove1993}.
-Crete is home to the only endemic mammal of Greece, the Cretan shrew (\textit{Crocidura zimmermanni}), more than 350 endemic arhropods, \# endemic plants among them a tree \textit{Zelkova abelicea}.
-Crete is a continental island, fifth largest island of the mediterranean, a mediterranean biodiversity hotspot, model for desertification, climate change \cite{Kougioumoutzis2020} and habitat loss \cite{ISPIKOUDIS1993259},
-elevation gradient \cite{trigas2013, FAZAN2017}, grasing impacts on ecosystems \cite{JouffroyBapicot2016} and evolutionary model \cite{POULAKAKIS2002}.
-Yet the terrestrial microbial diversity has been unexplored.
+Crete is a continental forarc island \cite{ali2016}, fifth largest island of the mediterranean (~8300 km\textsuperscript{2}),
+and a mediterranean biodiversity hotspot \cite{myers2000biodiversity}.
+The island of Crete has been studied since the classical times for its'
+fauna \cite{Sidiropoulos_Polymeni_Legakis_2017,Anastasiou2018Tenebrionid}, flora \cite{Krimbas_2005} and ecosystems \cite{Grove1993}.
+Crete is home to the only endemic mammal of Greece, the Cretan shrew (\textit{Crocidura zimmermanni}),
+more than 350 endemic arthropods \cite{bolanakis2024} and 183 endemic plants \cite{Kougioumoutzis2020}
+among them a tree \textit{Zelkova abelicea}. Multifacet factors have shaped the
+biodiversity of the island, for example the sharp elevation gradient \cite{trigas2013, FAZAN2017},
+the complex evolutionary history \cite{POULAKAKIS2002} and the human - nature
+interactions over thousands of years \cite{Vogiatzakis2008_med, Sfenthourakis2017}.
+The major threats of human activities are becoming apperend in the island's ecosystems,
+like desertification \cite{KARAMESOUTI2018266}, intesive grasing \cite{JouffroyBapicot2016},
+climate change \cite{Kougioumoutzis2020,Vogiatzakis2016} and habitat loss \cite{ISPIKOUDIS1993259}.
+Yet the topsoil microbial diversity of Crete has been unexplored.
 
-Has been sampled from major worldwide projects of microbiome studies \cite{Vasar2022, Labouyrie2023, Bahram2018, Orgiazzi2018}.
-Soil fungi \cite{Mikryukov2023, Davison2021, Tedersoo2021} and other soil eukaryotes \cite{Aslani2022}
+Worldwide projects of microbiome studies have collected topsoil samples from Crete \cite{Vasar2022, Labouyrie2023, Bahram2018, Orgiazzi2018}.
+Some have focused on soil fungi \cite{Mikryukov2023, Davison2021, Tedersoo2021} and other soil eukaryotes \cite{Aslani2022}.
+These studies have
 
-\cite{Vogiatzakis2008_crete}
-\cite{Vogiatzakis2016}
+Koiliaris
 
-\cite{Sfenthourakis2017}
-\cite{Vogiatzakis2008_med}
+Soil structure and chemistry, holistic
+
+Spatial data
 

### 2024-04-04 manuscript.tex
platforms and soil data

@@
 defies their macroscopic functionality.
-Global soil microbiome studies have been employed to decipher soil microbiome compositions \cite{Delgado-Baquerizo2018, Labouyrie2023},
+Global soil microbiome studies have been employed to decipher soil microbiome
+compositions \cite{thompson2017a-communal, Delgado-Baquerizo2018, Labouyrie2023},
 functions \cite{Bahram2018} and biogeography \cite{Martiny2006, Guerra2020}.
@@
 in a soil microbiome survey with a dense sampling, 0.017 samples per km\textsuperscript{2}.
-Hence, ISD is large scale study in the confined space of the island of Crete which 
+Hence, ISD is a large scale study in the confined space of the island of Crete which 
 is considered a miniature continent \cite{Vogiatzakis2008_crete}.
@@
 
-Worldwide projects of microbiome studies have collected topsoil samples from Crete \cite{Vasar2022, Labouyrie2023, Bahram2018, Orgiazzi2018}.
-Some have focused on soil fungi \cite{Mikryukov2023, Davison2021, Tedersoo2021} and other soil eukaryotes \cite{Aslani2022}.
-These studies have
+Worldwide projects of microbiome studies have collected one or two topsoil
+samples from Crete \cite{Vasar2022, Labouyrie2023, Bahram2018, Orgiazzi2018}.
+Some have focused on soil fungi \cite{Mikryukov2023, Davison2021, Tedersoo2021}
+and other soil eukaryotes \cite{Aslani2022}.
+The only thorough microbiome study of a soil ecosystem in Crete, to our knowledge,
+is in the north west part of the island, the Koiliaris Critical Zone Observatory \cite{tsiknia2014}.
+Apart from the soil microbiome, soil physical and chemical properties has been
+investigaded by globan and european projects like the Forum of European Geological Surveys
+(FOREGS) \cite{nerc19017}, the Geochemical Mapping of Agricultural and Grazing Land
+Soil in Europe (GEMAS) \cite{REIMANN2018302} and the Soil Profile Analytical
+Database for Europe (SPADE) \cite{Hiederer2006}.
 
-Koiliaris
+Data and metadata of these samplings are stored in different databases, yet 
+great effort from these distinct communities have led to establishing standards
+to enable FAIR data \cite{Wilkinson2016}. For amplicon sequences the Genome Standards
+Consortium \cite{Field2011} has established the MIMARKS \cite{yilmaz2011minimum}
+standards and has been a advocate for open and unrestricted data \cite{Amann2019}.
+For soil data, rich metadata and platforms of hosting open data exist, e.g the 
+European Soil Data Centre \cite{Panagos2022} and the World Soil Information
+Service (WoSIS) \cite{Batjes2024}. These platforms also provide maps.
+
+Spatial and remote sensing data
 
@@
 
-Spatial data
+Integration of prior knowledge and
 
-Integration of prior knowledge and 
-Genome Standards Consortium \cite{Field2011}, FAIR data \cite{Wilkinson2016}, unrestricted \cite{Amann2019},
-rich metadata and platforms of hosting open data like European Soil Data Centre \cite{Panagos2022}
+Interactomes have been characterised as a driver of community composition.
+Microbial interactions challenges \cite{Faust2021} cohesion?
 
-Interactomes as a driver of community composition. Microbial interactions challenges \cite{Faust2021} cohesion?
-
-In this study, we investigate
+In this study, we integrade multiple types of data and methods to decipher hidden 
+signals of the topsoil bacterial communities of Crete.
 

### 2024-04-04 manuscript.tex
add standards and integration in intro

@@
 
-% \boxedtext{
-% \begin{itemize}
-% \item Key boxed text here.
-% \item Key boxed text here.
-% \item Key boxed text here.
-% \end{itemize}}
-
 \maketitle
@@
 
-Crete is a continental forarc island \cite{ali2016}, fifth largest island of the mediterranean (~8300 km\textsuperscript{2}),
+Crete is a continental forarc island \cite{ali2016}, fifth largest island of the mediterranean (8350 km\textsuperscript{2}),
 and a mediterranean biodiversity hotspot \cite{myers2000biodiversity}.
@@
 Consortium \cite{Field2011} has established the MIMARKS \cite{yilmaz2011minimum}
-standards and has been a advocate for open and unrestricted data \cite{Amann2019}.
-For soil data, rich metadata and platforms of hosting open data exist, e.g the 
+standards among others, and has been a advocate for open and unrestricted data \cite{Amann2019}.
+Examples of rich metadata and platforms of hosting open data and digital soil maps are the 
 European Soil Data Centre \cite{Panagos2022} and the World Soil Information
-Service (WoSIS) \cite{Batjes2024}. These platforms also provide maps.
+Service (WoSIS) of the ISRIC \cite{Batjes2024}. Apart from samplings,
+spatial data are openly available terrestrial ecosystems.
+Climatic data, land cover, desertification risk, aridity, soil type, normalized
+vegetation index, bedrock geological formations. From the bacterial point of view, 
+there curated databeses that classify in some baseline functionality.
 
-Spatial and remote sensing data
+Soil ecosystem functioning
 
-Soil structure and chemistry, holistic
+Interactomes have been characterised as a driver of community composition soil
+functioning \cite{GUSEVA2022108604}.
+Still the challenges to infer actual microbial interactions remain \cite{Faust2021}.
 
@@
 
-Interactomes have been characterised as a driver of community composition.
-Microbial interactions challenges \cite{Faust2021} cohesion?
+UN and EU soil goals for the 2030
 
-In this study, we integrade multiple types of data and methods to decipher hidden 
-signals of the topsoil bacterial communities of Crete.
+In this study we ask: what the differences in the microbiome communities in different land use types?
+Are there any climatic, geological, elevational, aridity or functional factors that affected the differences?
+How the interactome changes over different land use types and climate?
+Are there any distictions between arid regions of Crete?
+To address these questions, we integrate multiple types of data and methods to decipher hidden 
+signals of the Island Sampling Day soil microbiome data. In total, 144 samples from
+72 sites (2 samples per site) of Crete are used with their metadata.
+Warning signals are also identified for various types of ecosystems.
 
@@
 processes and interactions of fauna, flora and microbes \cite{Fry2019, Crowther2019,GRANDY201640,Delgado-Baquerizo2020}
+Positive and negative in soils \cite{Liu2024}
 
@@
 and soil conservation instead of specific species \cite{Guerra2021}
-and policy \cite{KONINGER2022} across countries \cite{Putten2023}
+and policy \cite{KONINGER2022} across countries \cite{Putten2023} and in Greece \cite{SCHISMENOS2022100035}
 
-"A holistic perspective on soil architecture is needed as a key to soil functions"
+"A holistic perspective on soil architecture is needed as a key to soil functions" \cite{philippot2024the-interplay}
 \section{Conclusion}

### 2024-04-04 manuscript.tex
intro first draft ready

@@
 
-Soil ecosystem functioning
-
-Interactomes have been characterised as a driver of community composition soil
+Disentagling the soil ecosystem functioning requires an holistic and 
+multidisciplinary approach \cite{vogel2022}. The integration of the afformentioned
+data is needed to understand the biogeochemical cycles along with biodiversity interactomes 
+that have been characterised as a driver of community composition soil
 functioning \cite{GUSEVA2022108604}.
-Still the challenges to infer actual microbial interactions remain \cite{Faust2021}.
-
-Integration of prior knowledge and
-
-UN and EU soil goals for the 2030
+Regarding the latter there are still big challenges to infer actual microbial
+interactions remain \cite{Faust2021}. All this work is needed in order to meet
+UN and EU soil goals for the 2030 and 2050 for healthy soils \cite{LAL2021e00398}.
 

### 2024-04-05 manuscript.tex
add workflow figure and some methods

@@
 They infuence and are infuenced by their environment and their community structure 
-defies their macroscopic functionality.
+defies their macroscopic functionality \cite{Bahram2018}.
 Global soil microbiome studies have been employed to decipher soil microbiome
@@
 study is by \cite{Karimi2020} which exemplified the 
-vast complexity of soil bacterial communuties and the requirement of dense samplings and isolated systems.
+vast complexity of soil bacterial communuties and the requirement of
+dense samplings and isolated systems \cite{Dini-Andreote2021}.
 
@@
 Borrowing this paradigm, soil microbes \cite{Li2020} and mycorrhizal fungi \cite{Delavaux2021} studies
-in islands show the benefits of using islands as models. Ultimately islands canr
+in islands show the benefits of using islands as models. Ultimately islands can
 set the ground to represent ecosystems with data and complex interactions \cite{Davies2016}.
@@
 
+\begin{figure}[t] 
+    \centering\includegraphics[width=\columnwidth]{crete_soil_microbiome}
+    \caption{Workflow of this study.
+Data integration of ISD data with multiple types of spatial data.
+Then, a threefold analysis of function annotations, network analysis and differential abundance.
+All these data and methods are used to focus on specific taxa, ecosystems and threats.}
+    \label{fig:workflow}
+\end{figure}
+
 \subsection{Island Sampling Day: Crete}\label{isd_data}
 
-The data for this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB21776
-ENA API \cite{Yuan2023}
+This study uses the ISD Crete data that have been deposited
+in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB21776.
+The raw sequences (fastq files) and metadata (xml files) were downloaded with custom scripts using the ENA API \cite{Yuan2023}.
+Specific information regarding the ISD sampling, DNA extractions, PCR and sequencing
+are discussed here \cite{holm2024}.
 
-\subsection{Data integration}\label{data}
+Amplicon Sequence Variants were inferred using DADA2 \cite{Callahan2016} for 
+filtering, denoising and chimeric reads removal. Normalisation of the reads
+across samples was implemented with the SRS R package \cite{Beule2020}. Samples
+with less than 10000 reads were removed. ASVs were assigned to taxonomy using 
+DADA2 with Silva 138 \cite{Quast2012}.
 
-Copernicus CORINE Land Cover \cite{CLC2023}
-WorldClim 2.0 \cite{Fick2017}
-World Database on Protected Areas (WDPA) \cite{Hanson2022}
-Function annotation - FAPROTAX \cite{Louca2016}
-Crete geology from IGME
-Aridity
-%NDVI
-Taxonomy Silva 138 \cite{Quast2012}
+\subsection{Crete data cube}\label{data}
+
+The complilation of Crete data cube has multiple spatial data layers of global,
+european, greek or cretan scale. 
+The Copernicus CORINE Land Cover has 3 layers resolution that classify the land
+use and cover in shapefile format \cite{CLC2023}. 
+WorldClim 2.0 contains global climatic data for 12 variables, e.g annual mean
+temperature and annual precipitation \cite{Fick2017}.
+We also utilised teh Environmentally Sensitive Areas Index to desertification (ESAI) 
+of Greece dataset \cite{KARAMESOUTI2018266}. Additionally we included the 
+Global Aridity Index and Potential Evapotranspiration Database \cite{zomer2022version}.
+Crete geology, bedrock formations were downloaded from IGME.
+Handling and analysis of these data was done with the sf and stars R packages \cite{Pebesma2023}.
+%World Database on Protected Areas (WDPA) \cite{Hanson2022}
+%Average NDVI, 2000-2020
+%Harmonized World Soil Database
+\subsection{Network inferrence}\label{network_inferrence}
+
+Network inferrence using FlashWeave 0.19.2 \cite{Tackmann2019}
+Network analysis in igraph R package \cite{Csardi2006}
+Julia Version 1.9.3 \cite{Julia-2017}
 
 \subsection{Tools}\label{tools}
-dada2 for ASV inferrence \cite{Callahan2016}
-PEMA for OTU inferrence \cite{Zafeiropoulos2020}
-ape R packege \cite{Paradis2004}
-SRS R package \cite{Beule2020}
+Function annotation - FAPROTAX \cite{Louca2016}
+%PEMA for OTU inferrence \cite{Zafeiropoulos2020}
+%ape R packege \cite{Paradis2004}
 UMAP ordination \cite{mcinnes2018umap-software}
@@
 Python 3.11.4
-Julia Version 1.9.3 \cite{Julia-2017}
 R version 4.3.2 \cite{rcoreteam}
-Network inferrence using FlashWeave 0.19.2 \cite{Tackmann2019}
-Network analysis in igraph R package \cite{Csardi2006}
 Differential abundance analysis with ANCOM-BC R package \cite{Lin2020,Lin2023}
-Spatial analysis with the sf and stars R packages \cite{Pebesma2023}
 U-CIE R package for coloring 3 dimentional data \cite{Koutrouli2022}
@@
 %\end{appendices}
+\section{Data and Code}
+The documentation and scripts developed for this study are available in
+\href{https://github.com/savvas-paragkamian/crete_soil_microbiome/}{Crete soil microbiome github repository}.
+This repository contains all the necessary scripts for the data retrieval,
+filtering and ASV inferrence, taxonomy assignment, data integration of spatial data, 
+functional annotation and the subsequent analyses and visualisation.
+Code is structured to be reproducible and interoperable.
 
@@
 \section{Author contributions statement}
-
-Must include all authors, identified by initials, for example:
-S.R. and D.A. conceived the experiment(s),  S.R. conducted the experiment(s), S.R. and D.A. analysed the results.  S.R. and D.A. wrote and reviewed the manuscript.
+Conceptualization: LS, EP, GK, AM;
+Data curation: JH, LS, EP, SY, SP;
+Formal Analysis: SP, JH;
+Funding acquisition: LS, AM, SP;
+Investigation: LS, SY, SP, JH;
+Methodology: SP, GK, EP, LS, SY, CC, CP, HZ, MS;
+Project administration: LS, EP, GK;
+Resources: EP, LS, SY, AM;
+Software: SP, JH, HZ;
+Supervision: LS, EP, GK, PS;
+Validation: JH, CP, CC, HZ, PS, DT, MS, LS;
+Visualization: SP, JH;
+Writing – original draft: SP, EP, GK;
+Writing – review and editing: All authors.
 
 \section{Acknowledgments}
-The authors thank the anonymous reviewers for their valuable suggestions. This work is supported in part by funds from the National Science Foundation (NSF: \# 1636933 and \# 1920920).
+The authors thank the anonymous reviewers for their valuable suggestions.
+
+\section{Funding}
+SP was supported by the 3rd H.F.R.I. Scholarships for PHD Candidates (no. 5726) for his work.
 

### 2024-04-06 manuscript.tex
update abstract to 250 words

@@
 
-\abstract{Ecosystem functioning is an integral part of the sciences of climate
-    change and conservation. Microbes are known for their versatility, abundance
-    and influence on ecosystem functions, but a more thorough analysis of
-    ecosystems should include a more complex assembly of organisms.
-    For example, plant-arthropod-soil microbiome interactions are important
-    associations that have historically been ignored. A synthesized knowledge
-    base of biodiversity, in terms of ecological and remote-sensing data remains a
-    major challenge. Many worldwide studies have been published regarding soil
-    microbiome ecosystems, though there are many gaps to cover the complexity
-    of functioning and biodiversity in the local scale. Islands can be
-    important case studies for this integration, from micro to macro, with Crete being a great example. 
+\abstract{
+    Microbes are known for their versatility, abundance
+    and influence on soil ecosystem functioning.
+    A synthesized knowledge base of microbial biodiversity, in terms of
+    ecological and remote-sensing data remains a major challenge.
+    Many worldwide studies have been published regarding soil
+    microbiome ecosystems, though there are still many blind spots.
+    Islands can be important case studies for this integration for more resolute and dense samplings.
     Here, we utilize the Island Sampling Day Crete 2016 microbial 16S rRNA gene
-    amplicon data and metadata, integrated with soil, faunistic and remote
+    amplicon data, integrated with soil and remote
     sensing data, to decipher the drivers of ecosystem function of the island.
-    Cretan macroecology has been studied for centuries for its diverse and
-    unique geology, fauna and flora. In addition, Crete as a continental island,
-    presents a distinct natural and evolutionary history with high contrasts in
-    vegetation cover, climatic conditions and geology. The Island Sampling Day
-    Crete 2016 project (co-hosted by the Genomic Standards Consortium (GSC) and
-    the Institute of Marine Biology, Biotechnology and Aquaculture (IMBBC-HCMR)), has
-    collected microbial 16s amplicon data from 72 distinct, with ecosystem
-    diversity topsoil sites from all around Crete, accompanied by FAIR
+    The Island Sampling Day Crete 2016 project has collected 144 topsoil samples
+    from 72 sites, capturing a lot of this diversity, accompanied by FAIR
     (Findable, Accessible, Interoperable and Reproducible) data by design. 
-    With this island wide study the GSC put the genomic standards in action.
-The preliminary results indicate a notable influence of the pH and the elevation
-over the island's microbiota. In particular, sites in higher altitudes found to
+    Cretan macroecology has been studied for centuries for its diverse  and endemic
+    fauna and flora.
+    In addition, Crete has been considered as a miniature continent with high contrasts in
+    vegetation cover, elevation, climatic conditions. 
+    We show that, higher altitudes in Crete found to
     be inhabited by a more diverse number of microorganisms, a pattern commonly
-seen in several faunistic groups, such as arthropods. These data along with the
-open access data regarding arthropod fauna, flora and the vast data of remote
-sensing and biodiversity, provide the basis to identify major drivers of
-biodiversity, to evaluate hotspots and contribute to foreknowledge of threatened ecosystems.}
+    seen in several faunistic groups, such as arthropods.
+    The integration of the spatial data presented here enabled warning signals
+    in pristine and grazing ecosystems.
+    These results along with the
+    climatic and desertification index influences on the soil microbiome of Crete,
+    provide the basis to identify major drivers of biodiversity, to evaluate hotspots
+    and contribute to foreknowledge of threatened ecosystems.
+}
 
@@
 Global Aridity Index and Potential Evapotranspiration Database \cite{zomer2022version}.
-Crete geology, bedrock formations were downloaded from IGME.
+Crete geological bedrock formations were downloaded from IGME.
 Handling and analysis of these data was done with the sf and stars R packages \cite{Pebesma2023}.
@@
 
-Network inferrence using FlashWeave 0.19.2 \cite{Tackmann2019}
-Network analysis in igraph R package \cite{Csardi2006}
-Julia Version 1.9.3 \cite{Julia-2017}
+The network inferrence was facilitated with FlashWeave 0.19.2 \cite{Tackmann2019}
+in Julia language version 1.9.3 \cite{Julia-2017}. To use FlashWeave we reduced 
+the abandunce table of the ASVs to keep the ones at the genus or species level.
+In addition, we filtered these taxa that appeared in ### samples and had more than 
+#### mean relative abundance. The subsequent network analyses were carried out
+with the igraph R package \cite{Csardi2006}.
 
 \subsection{Tools}\label{tools}
-Function annotation - FAPROTAX \cite{Louca2016}
 %PEMA for OTU inferrence \cite{Zafeiropoulos2020}
-%ape R packege \cite{Paradis2004}
-UMAP ordination \cite{mcinnes2018umap-software}
-Numerical ecology analyses with vegan R package \cite{vegan}
+%U-CIE R package for coloring 3 dimentional data \cite{Koutrouli2022}
 
-Python 3.11.4
-R version 4.3.2 \cite{rcoreteam}
-Differential abundance analysis with ANCOM-BC R package \cite{Lin2020,Lin2023}
-U-CIE R package for coloring 3 dimentional data \cite{Koutrouli2022}
-ggplot2 \cite{ggplot22016}
-pheatmap \cite{Kolde2019}
-
-Computations were performed on HPC infrastrure of HCMR \cite{Zafeiropoulos2021}.
+For taxa function annotation we used the manually curated FAPROTAX database and python script \cite{Louca2016}.
+Whereas for the differential abundance analysis with ANCOM-BC2 R package \cite{Lin2023}.
+Numerical ecology analyses,e.g diversity indices, NMDS, PERMANOVA we calculated
+with the vegan R package \cite{vegan}.
+For PCoA ordination we used ape R packege \cite{Paradis2004} and UMAP python library\cite{mcinnes2018umap-software}.
+Visualisation was implemented with ggplot2 \cite{ggplot22016} and pheatmap \cite{Kolde2019}.
+The environment we worked had Python 3.11.4 and R version 4.3.2 \cite{rcoreteam}. 
+Finally, computations were performed on HPC infrastrure of HCMR \cite{Zafeiropoulos2021}.
 

### 2024-04-06 manuscript.tex
update keywords

@@
     seen in several faunistic groups, such as arthropods.
-    The integration of the spatial data presented here enabled warning signals
+    The integration of the spatial data with state of the art methods enabled warning signals
     in pristine and grazing ecosystems.
@@
 
-\keywords{soil microbiome, networks, island biogeography, Crete, data integration, FAIR data}
+\keywords{soil microbiome, networks, island biogeography, Crete, spatial data, remote sensing, data integration, FAIR data}
 

### 2024-04-11 manuscript.tex
add appendix table

@@
 Specific information regarding the ISD sampling, DNA extractions, PCR and sequencing
-are discussed here \cite{holm2024}.
+protocols are discussed here \cite{holm2024}.
 
@@
 Global Aridity Index and Potential Evapotranspiration Database \cite{zomer2022version}.
-Crete geological bedrock formations were downloaded from IGME.
-Handling and analysis of these data was done with the sf and stars R packages \cite{Pebesma2023}.
+Geological formations shapefiles were downloaded from the geoportal of
+Decentralized Administration of Crete, which were developed by
+Crinno-Emeric Group project\footnote{\url{https://geoportal.apdkritis.gov.gr/gis/apps/storymaps/stories/19690f65abbe4e8ab0141b2fe7261a8c}}.
+Handling and analysis of these data was done with the sf and terra R packages \cite{Pebesma2023}.
 %World Database on Protected Areas (WDPA) \cite{Hanson2022}
-%Average NDVI, 2000-2020
 %Harmonized World Soil Database
-\subsection{Network inferrence}\label{network_inferrence}
 
-The network inferrence was facilitated with FlashWeave 0.19.2 \cite{Tackmann2019}
-in Julia language version 1.9.3 \cite{Julia-2017}. To use FlashWeave we reduced 
+\subsection{Integrative Analysis and Annotations}\label{int_analysis}
+The network inferrence was facilitated with FlashWeave 0.19.2 \cite{Tackmann2019}.
+To use FlashWeave we reduced 
 the abandunce table of the ASVs to keep the ones at the genus or species level.
-In addition, we filtered these taxa that appeared in ### samples and had more than 
-#### mean relative abundance. The subsequent network analyses were carried out
+In addition, we filtered these taxa that appeared in SOSO samples and had more than 
+SOSO mean relative abundance. The subsequent network analyses were carried out
 with the igraph R package \cite{Csardi2006}.
-
-\subsection{Tools}\label{tools}
-%PEMA for OTU inferrence \cite{Zafeiropoulos2020}
-%U-CIE R package for coloring 3 dimentional data \cite{Koutrouli2022}
 
@@
 For PCoA ordination we used ape R packege \cite{Paradis2004} and UMAP python library\cite{mcinnes2018umap-software}.
+
+\subsection{Tools}\label{Coding environment}
+%PEMA for OTU inferrence \cite{Zafeiropoulos2020}
+%U-CIE R package for coloring 3 dimentional data \cite{Koutrouli2022}
+
 Visualisation was implemented with ggplot2 \cite{ggplot22016} and pheatmap \cite{Kolde2019}.
-The environment we worked had Python 3.11.4 and R version 4.3.2 \cite{rcoreteam}. 
+The environment we worked had Python 3.11.4, R version 4.3.2 \cite{rcoreteam}
+and Julia language version 1.9.3 \cite{Julia-2017}in Julia language version 1.9.3 \cite{Julia-2017}.
 Finally, computations were performed on HPC infrastrure of HCMR \cite{Zafeiropoulos2021}.
@@
 
+Crete's ecosystems are mostly semi-arid, whereas in the mountain ranges there 
+are areas classified as dry sub-humid and humid.
+
+Parent material and infuences on soil functions and bacterial communities.
+
 Networks specialists generalists \cite{Barberan2012}
@@
 
-Future works, shotgun metagenomics, long reads, need for even higher resolution samplings.
+The pillar of data integration is the unrestricted open data across disciplines and 
+the open source software.
+Deciphering and validating the results presented here requires future work.
+Shotgun metagenomics and metatranscriptomics can uleash the functional potential of
+topsoil along with other advancements like long reads sequencing. Higher resolution
+samplings using grid system will enhance the resolution and also the resampling of
+ISD sites in different time points will provide additional insights to the complex soil 
+functions.
 
@@
 
-%\begin{appendices}
-%
-%\section{Appendix 1}\label{appendix1}
-%
-%Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
-%pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
-%vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
-%a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
-%urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris.
-%
-%
+\begin{appendices}
+
+\section{Appendix 1}\label{appendix1}
+
+Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
+pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
+vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
+a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
+urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris.
+
+\begin{table}[]
+    \caption{Taxonomic depth of ASVs and the unique number of taxa of each level.\label{table:asv_taxonomy}}%
+\begin{tabular}{@{}lll@{}}
+classification depth & Total ASV    & Total taxa \\
+Kingdom              & 1974         & 2          \\
+Phylum               & 4034         & 33         \\
+Order                & 38517        & 193        \\
+Class                & 24157        & 83         \\
+Family               & 71355        & 287        \\
+Genus                & 90137        & 1166       \\
+Species              & 9120         & 1338       \\
+Total                & $\sim$239000 & 3102      
+\end{tabular}
+\label{table:asv_taxonomy}
+\end{table}
+
 %\section{Appendix 2}\label{appendix2}
@@
 %
-%\end{appendices}
+\end{appendices}
 \section{Data and Code}

### 2024-04-13 manuscript.tex
table 1 summary of data cube ready

@@
 
+\begin{sidewaystable*}
+    \caption{Summary of the different spatial layers in Crete in terms of total area, number of samples and microbial diversity.\label{table:data_cube_summary}}
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}llllllll@{\extracolsep{\fill}}}
+\tabcolsep=0pt%
+class                                           & area & category             & samples & taxa richness & asv richness & mean shannon & sd shannon \\
+Arable land                                     & 88   & CLC LABEL2           & 4       & 1518           & 6178          & 4.73          & 0.17        \\
+Artificial, non-agricultural vegetated areas    & 21   & CLC LABEL2           & NA      & NA             & NA            & NA            & NA          \\
+Forests                                         & 300  & CLC LABEL2           & 4       & 1803           & 8663          & 4.91          & 0.18        \\
+Heterogeneous agricultural areas                & 1103 & CLC LABEL2           & 31      & 13701          & 59581         & 4.83          & 0.24        \\
+Industrial, commercial and transport units      & 40   & CLC LABEL2           & 4       & 1760           & 6517          & 4.79          & 0.32        \\
+Inland waters                                   & 7    & CLC LABEL2           & NA      & NA             & NA            & NA            & NA          \\
+Mine, dump and construction sites               & 10   & CLC LABEL2           & NA      & NA             & NA            & NA            & NA          \\
+Open spaces with little or no vegetation        & 411  & CLC LABEL2           & 7       & 3132           & 14568         & 4.85          & 0.21        \\
+Pastures                                        & 59   & CLC LABEL2           & 4       & 1609           & 7535          & 4.85          & 0.14        \\
+Permanent crops                                 & 2368 & CLC LABEL2           & 22      & 9999           & 43833         & 4.88          & 0.18        \\
+Scrub and/or herbaceous vegetation associations & 3798 & CLC LABEL2           & 58      & 24394          & 110127        & 4.76          & 0.29        \\
+Urban fabric                                    & 111  & CLC LABEL2           & 4       & 1765           & 9828          & 4.78          & 0.19        \\
+-                                               & 1    & Geology              & NA      & NA             & NA            & NA            & NA          \\
+J-E                                             & 1347 & Geology              & 22      & 8995           & 41730         & 4.75          & 0.25        \\
+K-E                                             & 248  & Geology              & NA      & NA             & NA            & NA            & NA          \\
+K.k                                             & 1253 & Geology              & 27      & 11277          & 51737         & 4.72          & 0.25        \\
+K.m                                             & 13   & Geology              & NA      & NA             & NA            & NA            & NA          \\
+Mk                                              & 812  & Geology              & 13      & 5618           & 24633         & 4.79          & 0.2         \\
+Mm.I                                            & 1614 & Geology              & 12      & 5259           & 22854         & 4.86          & 0.15        \\
+Ph-T                                            & 1012 & Geology              & 14      & 6720           & 29742         & 4.92          & 0.19        \\
+Q.al                                            & 911  & Geology              & 34      & 15580          & 64851         & 4.91          & 0.27        \\
+T.br                                            & 300  & Geology              & 2       & 625            & 2839          & 4.4           & 0.24        \\
+f                                               & 118  & Geology              & 2       & 720            & 4753          & 4.5           & 0.01        \\
+fo                                              & 318  & Geology              & 2       & 825            & 4606          & 4.65          & 0.04        \\
+ft                                              & 276  & Geology              & 10      & 4062           & 19085         & 4.79          & 0.17        \\
+o                                               & 94   & Geology              & NA      & NA             & NA            & NA            & NA          \\
+N                                               & 163  & Desertification Risk & NA      & NA             & NA            & NA            & NA          \\
+F2                                              & 1945 & Desertification Risk & 53      & 23404          & 103956        & 4.82          & 0.26        \\
+F1                                              & 1518 & Desertification Risk & 22      & 8817           & 41020         & 4.7           & 0.23        \\
+P                                               & 1593 & Desertification Risk & 36      & 16256          & 73256         & 4.86          & 0.19        \\
+C2                                              & 638  & Desertification Risk & 6       & 2113           & 9105          & 4.59          & 0.22        \\
+Other areas                                     & 290  & Desertification Risk & NA      & NA             & NA            & NA            & NA          \\
+F3                                              & 1144 & Desertification Risk & 14      & 5900           & 25362         & 4.82          & 0.28        \\
+C1                                              & 788  & Desertification Risk & 7       & 3191           & 14131         & 4.93          & 0.25        \\
+C3                                              & 238  & Desertification Risk & NA      & NA             & NA            & NA            & NA          \\
+Semi-Arid                                       & 6336 & Aridity class        & 98      & 43064          & 184930        & 4.83          & 0.25        \\
+Dry sub-humid                                   & 1343 & Aridity class        & 32      & 13649          & 65134         & 4.79          & 0.21        \\
+Humid                                           & 545  & Aridity class        & 8       & 2968           & 16766         & 4.55          & 0.19       
+\end{tabular*}
+\end{sidewaystable*}
+
 \subsection{Community drivers}\label{Drivers}
@@
 \begin{appendices}
+\setcounter{table}{0}
+\renewcommand{\thetable}{A\arabic{table}}
 
 \section{Appendix 1}\label{appendix1}
-
-Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
-pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
-vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
-a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
-urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris.
 
@@
 Visualization: SP, JH;
-Writing – original draft: SP, EP, GK;
+Writing – original draft: SP;
 Writing – review and editing: All authors.

### 2025-01-26 manuscript.tex
authors list
update affiliations

@@
 
-\title[Crete's soil microbiome]{Exploring the Soil Microbiome coupling with the Crete island Data Cube}
+\title[Crete's soil microbiome]{Exploring the Island Soil Microbiome coupling with spatial data cube}
 
@@
 \author[3,4]{Johanna B. Holm\ORCID{0000-0002-7646-4085}}
+\author[6]{Haris Zafeiropoulos\ORCID{0000-0002-4405-6802}}
 \author[5]{Christina Pavloudi\ORCID{0000-0001-5106-6067}}
-\author[6]{Haris Zafeiropoulos\ORCID{0000-0002-4405-6802}}
-\author[1]{Christos A. Christakis\ORCID{0000-0002-7075-0996}}
+\author[2]{Christos A. Christakis\ORCID{0000-0002-7075-0996}}
 \author[1]{Melanthia Stavroulaki\ORCID{0000-0003-4392-7159}}
-\author[1]{Dimitris Tsaparis\ORCID{0000-0001-9626-5553}}
-%\author[1]{Iraklis Vretzakis\ORCID{0000-0001-9626-5553}}
+%\author[1]{Dimitris Tsaparis\ORCID{0000-0001-9626-5553}}
+%\author[1]{Iraklis Vretzakis}
+\author[8]{Ioannis Antoniou\ORCID{0000-0001-8080-0945}}
+\author[2]{Panagiotis F. Sarris\ORCID{0000-0001-7000-8997}}
 \author[7]{Stephanie A. Yarwood\ORCID{0000-0003-4593-7908}}
-\author[2]{Antonios Magoulas}
-\author[1]{Panagiotis F. Sarris\ORCID{0000-0001-7000-8997}}
-\author[2]{Giorgos Kotoulas\ORCID{0000-0002-4667-8533}}
+\author[1]{Giorgos Kotoulas\ORCID{0000-0002-4667-8533}}
+\author[1]{Antonios Magoulas}
 \author[3]{Lynn Schriml\ORCID{0000-0001-8910-9851}}
-\author[2,$\ast$]{Evangelos Pafilis\ORCID{0000-0001-5079-0125}}
+\author[1,$\ast$]{Evangelos Pafilis\ORCID{0000-0001-5079-0125}}
 
@@
 
-\address[1]{\orgdiv{Department of Biology}, \orgname{University of Crete}, \orgaddress{\street{Voutes}, \postcode{70013}, \state{Crete}, \country{Greece}}}
-\address[2]{\orgdiv{Institute of Marine Biology, Biotechnology and Aquaculture (IMBBC)}, \orgname{Hellenic Centre for Marine Research (HCMR)}, \orgaddress{\street{Gournes}, \postcode{71500}, \state{Crete}, \country{Greece}}}
+\address[1]{\orgdiv{Institute of Marine Biology, Biotechnology and Aquaculture (IMBBC)}, \orgname{Hellenic Centre for Marine Research (HCMR)}, \orgaddress{\street{Gournes}, \postcode{71500}, \state{Crete}, \country{Greece}}}
+\address[2]{\orgdiv{Department of Biology}, \orgname{University of Crete}, \orgaddress{\street{Voutes}, \postcode{70013}, \state{Crete}, \country{Greece}}}
 \address[3]{\orgdiv{Institute for Genome Sciences}, \orgname{University of Maryland School of Medicine}, \orgaddress{\state{Baltimore}, \country{USA}}}
@@
 \address[7]{\orgdiv{Department of Environmental Science and Technology}, \orgname{University of Maryland}, \orgaddress{\state{Baltimore}, \country{USA}}}
+\address[8]{\orgdiv{Department of Mathematics}, \orgname{Aristotle University of Thessaloniki}, \orgaddress{\postcode{54124}, \state{Thessaloniki}, \country{Greece}}}
 
@@
 
+one of the first soil metagenomic studies \cite{fierer2012cross-biome}
+
+history of 16s \cite{}
+
+\cite{baykal2024genomic}
 Soil ecosystems are the cornerstone of terrestrial habitats, biodiversity and henceforth human activities.
@@
 \bibliographystyle{unsrt}
-\bibliography{reference}
+\bibliography{crete_isd_health_refs}
 
